### 2021-11-15 documents/sections/extensions.tex
revised proto-pekodian *k-

\section{Incomplete extensions: the innovative \gl{1}\texorpdfstring{\gl{s_a_}}{Sa} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the person marker extensions which did not affect all potential targets all have in common that they feature innovative first person markers on verbs that are (at least historically) members of the \gl{s_a_} class.
Of the six attested incomplete extensions, three can be reconstructed to intermediate proto-languages, while three others happened in earlier stages of single modern languages.
The sources of innovative markers vary, but not much: the innovative \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
\cref{sec:pekodian} investigates the innovation of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} takes a closer look at the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} concerns the extension of \rc{t-} in \PTir (or \PTar), reflected in modern \trio and \akuriyo.
\crefrange{sec:akuriyo}{sec:yukpa} look at innovative first person markers which are only attested in single modern languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

\subsection{\PPek \rc{k-}}
\label{sec:pekodian}
The Pekodian branch was suggested by \textcite{meira2005southern}, as the result of fieldwork on \bakairi by Meira and the availability of more material on \ikpeng. % \parencites{pacheco1997relativizacao}{ikpengpacheco1997}{campetela1997analise}{pacheco1998nominalizacao}{ikpengpacheco2001}{campetela2002aspectos}{pacheco2002aspectos}{pacheco2003intransitivos}.
It consists of closely related \arara and \ikpeng, with \bakairi as a more distant member.
\textcite{meira2005southern} focused on phonological and lexical properties, so no reconstructive work on \PPek morphosyntax can be found in the literature.
All three Pekodian languages have a regular \gl{1}\gl{s_a_} marker \obj{k-}, as evidenced by the paradigms in \cref{tab:pekreg}.
We can thus reconstruct a \PPek \gl{1}\gl{s_a_} marker \rc{k-}.

@@

For \arara, \textcite{alves2017arara} describes six\footnote{Seven under her analysis, which sees the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.} \gl{s_a_} verbs which have a first person marker \obj{w(ɨ)-} rather than \obj{k-}, listed in \exref{arairr}.
From a comparative view this list is not quite complete, as there is also a reflex of the copula \rc{a[p]}, serving syntactically as a postposition introducing adverbial clauses meaning \qu{if} or \qu{when} \parencite[199--201]{alves2017arara}.
However, its inflectional morphology features verbal \setone prefixes, including first person \obj{w-} \exref{araap}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

\begin{multicols}{2}
\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
\ex<araap> \arara \parencite[200]{alves2017arara}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-aptam} \qu{when/if I was}\\
\gl{2} & \obj{m-od-aptam}\\
\gl{1+2} & \obj{kud-aptam}\\
\gl{3} & ∅\obj{-aptam}\\
\end{tabular}
\xe
\end{multicols}

@@

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
Of course, \qu{to bathe} is also found in the \obj{w-}list for \arara.
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.%
\footnote{
It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}.
However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}.
Given the lack of data for other verbs, we will not further discuss this potential additional pattern.
If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.
}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them in \exref{bakverbs} according to our current understanding of \bakairi phonology and verbal morphology, based on \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi}{franchetto2016classes}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
Further, while there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} reflexes seem to occur with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The presence and distribution of the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø suggests that it is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}.
Indeed, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
While it is by no means a regular sound change, it allows us to securely connect the two prefixes.
Similarly, the supposed change of \rc{wɨ} to \bakairi \obj{u} is found in other correspondences, like \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
Thus, we reconstruct a \gl{1}\gl{s_a_} prefix \rc{w(ɨ)-} to \PPek, identical to the \arara one in form and distribution.

As for the forms of the verb stems, a few comments are necessary:
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.
For \qu{to go down}, we reconstruct \rc{ɨ} as the initial vowel rather than \rc{i}, see \cref{sec:godown}.
Further, the forms are not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
However, the addition of a final \obj{ŋ} in \PXin is attested elsewhere,%
\footnote{
\begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
and based on the fact \bakairi has generally lost much segmental material, we suggest that both \obj{ŋ} and \obj{gɨ} are later additions to a root \rc{ɨptə}.

@@

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
On the other hand, both \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following the line of reasoning used by \textcite[114]{meira1998proto} (see also \cref{sec:taranoan}) for a similar pattern in the three Taranoan languages, we suggest that the idiosyncratic pattern in \arara is reconstructible to \PPek, and that \bakairi and \ikpeng independently regularized the paradigm to only use \rc{əd-epɨ}.

The V-initial nature of \qu{to go} is evidenced in its Xinguan forms, while \bakairi \phon{\rc{wɨ}}{\obj{u}} muddied the morpheme boundary; other forms are V-initial \pcref{sec:go}.
The \ikpeng form \obj{aran} is compatible with our reconstruction \rc{ɨtən} when considering that \ikpeng \obj{a} is an attested outcome of \rc{ə}.\footnote{\begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.}
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
In later, individual developments, \bakairi extended \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.
Further, both languages regularized the paradigm of \qu{to come} to \rc{əd-epɨ}, accompanied by the extension of \obj{k-}.

>>>>

\section{Incomplete extensions: the innovative \gl{1}\texorpdfstring{\gl{s_a_}}{Sa} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the person marker extensions which did not affect all potential targets all have in common that they feature innovative first person markers on verbs that are (at least historically) members of the \gl{s_a_} class.
Of the six attested incomplete extensions, three can be reconstructed to intermediate proto-languages, while three others happened in earlier stages of single modern languages.
The sources of innovative markers vary, but not much: the innovative \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2}\gl{p}/\gl{s_p_} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{p}/\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
We discuss each extension separately, contrasting regular and innovative verbs with irregular and conservative verbs, and reconstructing forms where necessary:
\cref{sec:pekodian} investigates the innovation of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} takes a closer look at the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} concerns the extension of \rc{t-} in \PTir (or \PTar), reflected in modern \trio and \akuriyo.
\crefrange{sec:akuriyo}{sec:yukpa} look at innovative first person markers which are only attested in single modern languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

\subsection{\PPek \rc{k-}}
\label{sec:pekodian}
The Pekodian branch was suggested by \textcite{meira2005southern}, as the result of fieldwork on \bakairi by Meira and the availability of more material on \ikpeng. % \parencites{pacheco1997relativizacao}{ikpengpacheco1997}{campetela1997analise}{pacheco1998nominalizacao}{ikpengpacheco2001}{campetela2002aspectos}{pacheco2002aspectos}{pacheco2003intransitivos}.
It consists of closely related \arara and \ikpeng, with \bakairi as a more distant member.
\textcite{meira2005southern} focused on phonological and lexical properties, so no reconstructive work on \PPek morphosyntax can be found in the literature.
However, all three Pekodian languages have a regular \gl{1}\gl{s_a_} marker \obj{k-}, as evidenced by the paradigms in \cref{tab:pekreg}.
Thus, it is possible to reconstruct a \PPek \gl{1}\gl{s_a_} marker \rc{k-}.

@@

In the most detailed description of a Pekodian language, \textcite{alves2017arara} describes six%
\footnote{Seven under her analysis, which sees the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.}
\arara \gl{s_a_} verbs forming a subclass defined by a first person marker \obj{w(ɨ)-} rather than \obj{k-}, listed in \exref{arairr}.
From a comparative perspective, this list is not quite complete, as there is also a reflex of the copula \rc{a[p]}, serving syntactically as a postposition introducing adverbial clauses meaning \qu{if} or \qu{when} \parencite[199--201]{alves2017arara}.
However, its inflectional morphology features verbal \setone prefixes, including first person \obj{w-} \exref{araap}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

\begin{multicols}{2}
\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe

\ex<araap> \arara \parencite[200]{alves2017arara}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-aptam} \qu{when/if I was}\\
\gl{2} & \obj{m-od-aptam}\\
\gl{1+2} & \obj{kud-aptam}\\
\gl{3} & ∅\obj{-aptam}\\
\end{tabular}
\xe
\end{multicols}

@@

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
Since \qu{to bathe} is also found in the \obj{w-}list for \arara, other \bakairi cognates of these verbs are of interest.
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.%
\footnote{
It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}.
However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}.
Given the lack of data for other verbs, we will not further discuss this potential additional pattern.
If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.
}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them in \exref{bakverbs} according to our current understanding of \bakairi phonology and verbal morphology, based on \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi}{franchetto2016classes}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
Further, while there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, it seems that only reflexes of \rc{eti} occur with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The presence and distribution of the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø suggests that it is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}.
Indeed, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
While it is by no means a regular sound change, it allows us to securely connect the two prefixes.
Similarly, the supposed change of \rc{wɨ} to \bakairi \obj{u} is found in other correspondences, like \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
Thus, we reconstruct a \gl{1}\gl{s_a_} prefix \rc{w(ɨ)-} to \PPek, identical to the \arara one both in form and distribution.

As for the forms of the verb stems, a few comments are necessary:
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.
For \qu{to go down}, we reconstruct \rc{ɨ} as the initial vowel rather than \rc{i} \pcref{sec:godown}.
Further, the forms are not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
However, the addition of a final \obj{ŋ} in \PXin is attested elsewhere,%
\footnote{
\begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
and based on the fact \bakairi has generally lost much segmental material, we suggest that both \obj{ŋ} and \obj{gɨ} are later additions to a root \rc{ɨptə}.

@@

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
On the other hand, both \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following the line of reasoning used by \textcite[114]{meira1998proto} (see also \cref{sec:taranoan}) for a similar pattern in the three Taranoan languages, we suggest that the idiosyncratic pattern in \arara is reconstructible to \PPek, and that \bakairi and \ikpeng independently regularized the paradigm to only use \rc{əd-epɨ}; similar issues are found outside of Pekodian \pcref{sec:come}.

Finally, the V-initial nature of \PPek \qu{to go} is evidenced in its Xinguan forms; while the \bakairi change \phon{\rc{wɨ}}{\obj{u}} obscured the morpheme boundary, other forms are V-initial \pcref{sec:go}.
The \ikpeng form \obj{aran} is compatible with our reconstruction \rc{ɨtən} when considering that \ikpeng \obj{a} is an attested outcome of \rc{ə}.\footnote{\begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.}
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
In later, individual developments, \bakairi extended \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.
Further, both languages regularized the paradigm of \qu{to come} to \rc{əd-epɨ}, accompanied by the introduction of first person5 \obj{k-}.
